---
title: "Analysis Pipeline"
author: "Abhinav Motheram"
date: "19/1/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

Set up the workflow pipeline and options. We first load the `targets` package and remove the potentially outdated workflow.

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(targets)
library(tarchetypes)
library(tflow)
library(fnmate)

library(knitr)
library(rmarkdown)

tar_unscript()
```

```{targets globals, tar_globals = TRUE}

library(conflicted)
library(dotenv)

library(targets)
library(tarchetypes)
library(here)
library(purrr, quietly = TRUE)

## renv::install("an0505/ctisglobal")
library(ctisglobal)

library(tidyverse)
library(vroom)
library(janitor)
library(lubridate)
library(readxl)

library(here)

conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")


functions <- list.files(here("R"), full.names = TRUE)
walk(functions, source)
```

## Define Targets

## Data Import

### COVID-19 India Targets

```{targets c19_bharat_data}
tar_target(
  c19_bharat_data,
  get_covid_bharat_data(),
  cue = tarchetypes::tar_cue_age(
    name = c19_bharat_data,
    age = as.difftime(1, units = "days")
  )
)
```

### Population Targets

```{targets ind_pop_projections}
tar_target(
  ind_pop_projections,
  clean_pop_estimates(excel_path = "data/Population projection_2020.xlsx")
)
```

```{targets ind_pop_adult_projections}
tar_target(
  ind_pop_adult_projections,
  read_excel(here("data/pop_projection_adult_oct2021.xlsx")) %>% 
    clean_names() %>% 
    select(state, person, male, female) %>% 
    mutate_if(is.double, as.integer) %>% 
    mutate(across(c("person", "male", "female"), ~ .x * 1000)) %>%
    mutate(state = case_when(state == "India" ~ "All Regions",
                             state == "Delhi" ~ "NCT of Delhi",
                             TRUE ~ state)),
  format = "fst_tbl"
)
```

### CTIS Targets

```{targets ctis_public_doc}
tar_target(
  ctis_public_doc,
  ctis_get_public_apidoc(),
  format = "fst_tbl"
)
```

```{targets ctis_public_indicators}
tar_target(
  ctis_public_indicators,
  ctis_public_doc %>%
    filter(indicator %in% c("covid", "flu", "anosmia", "vaccine_acpt", "covid_vaccine", "twodoses")) %>%
    pull(indicator)
)

```

```{targets ctis_public_regions}
tar_target(
  ctis_public_regions,
  ctis_get_public_regions(),
  format = "fst_tbl"
)
```

```{targets ctis_public_regions_india}
tar_target(
  ctis_public_regions_india,
  ctis_public_regions %>%
    filter(country == "India") %>%
    pull(region) %>%
    append(NA)
)
```

```{targets ctis_india_public}
tar_target(
  ctis_india_public,
  ctis_get_public_data(indicator = ctis_public_indicators, 
                       region = ctis_public_regions_india, 
                       date_end = as.character(Sys.Date() - 3)),
  pattern = cross(ctis_public_indicators, ctis_public_regions_india),
  format = "fst_tbl",
  cue = tarchetypes::tar_cue_age(
    name = ctis_india_public,
    age = as.difftime(1, units = "days")
  )
)

```

## Data Cleaning

Merge covid19bharat vaccination data and CTIS vaccination estimates to form a single table.

```{targets c19_vaccine_merged}
tar_target(
  c19_vaccine_merged,
  merge_vaccine_estimates(c19_bharat_data, 
                          ind_pop_adult_projections,
                          ctis_india_public),
  format = "fst_tbl"
)

```

Merge covid19bharat COVID diagnostic data and CTIS CLI estimates to form a single table.

```{targets c19_covid_merged}
tar_target(
  c19_covid_merged,
  merge_covid_estimates(c19_bharat_data,
                        ind_pop_projections,
                        ctis_india_public),
  format = "fst_tbl"
)

```

## Render Report

```{targets ctis_report}
tar_render(
  ctis_report,
  here("reports/vaccination_analysis_draft_v4.Rmd"),
  output_format ="all"
)

```

# Run the Pipeline

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tar_make()
```

